The increasing cost of happiness

### 2021-10-27 src/draft-SSM2.Rmd
Final response to Reviewer 1. Closes issue #27

1. Central Clinical School, Faculty of Medicine and Health, University of Sydney, NSW, Australia
2. ARC Centre of Excellence for Children and Families over the Life Course
3. School of Economics, University of Technology, NSW, Australia
4. Institute of Labor Economics (IZA), Bonn, Germany

@@

|            |                                     |
|------------|-------------------------------------|
| Draft      | `r format(Sys.time(), '%d %B, %Y')` |
| Words      | 4704                                |
| Tables     | 0                                   |
| Figures    | 4                                   |

@@

A fundamental question for society is how much happiness does a dollar buy? The accepted view among economists and psychologists is that money and happiness increase together up to a point, after which there is little further gain from increasing income. The location of this change point was reportedly USD\$75K in 2008, which was substantially more than the median income in the same year (USD\$52K), indicating the happiness or wellbeing of the poorest majority of the population was tied to marginal changes in income while the wellbeing of a richer minority was not. Thus the location of the change point has implications for the distribution of wellbeing between the rich and poor. Inequalities in wellbeing are increasingly relevant due to the growing recognition that increasing income does not neccesarily lead to equal changes in wellbeing. In this paper we estimate the relationship between income and subjective wellbeing over a 19-year period focusing on whether this change point has shifted over time, and explore its implications for inequalities in wellbeing. To date there has been no investigation as to whether the change point has increased or decreased over time so we formally tested for the presence of a change point as well temporal changes in the relationship between income and affective wellbeing (happiness), and income and cognitive wellbeing (life satisfaction), using household economic data from Australia between 2001-2019. The results revealed the change point between affective wellbeing and income has increased over those 19 years faster than inflation (i.e., cost of living). This suggests that inequalities in income may be driving increasing _inequities_ in happiness between the rich and the poor, with implications for health and recent government policy-goals to monitor and improve wellbeing.

@@

@@

- We investigate how the relationship between income and subjective wellbeing (affective and cognitive) has changed over time, using a dataset with highly detailed measurements of each
- Nineteen years of data from a well-developed, economically stable G20 nation is examined (Australia), allowing a relatively undisturbed view of the changes without major economic shocks
- We are able to show for the first time that the income level at which affective wellbeing (happiness) is less responsive to income has risen since 2001, resulting in a greater inequality in the distribution of affective wellbeing between the rich and the poor
- Conversely the linear relationship between income and cognitive wellbeing (life-satisfaction) has remained relatively constant over this period
- The association between happiness and income is becoming more prevalent in the population this millennium, and likely to drive inequities in wellbeing across the economy

@@

Subjective wellbeing is not a unitary entity [@diener2017findings]; studies typically distinguish between life satisfaction, the cognitive appraisal of one’s own accomplishments, and affective wellbeing, one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. While measures of these distinct concepts are correlated, the distinction seems to be critical in understanding the impact of income on wellbeing [@howell2008relation]. For instance, we have recently observed that a major financial gain substantially improved cognitive wellbeing for individuals without much impact on their affective wellbeing [@kettlewell2020differential]. In a now classic paper, Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with log household income up to a point (USD75,000), but after that, increasing income had no further effect (i.e., "satiation"). Conversely cognitive wellbeing continued to increase with (log) income beyond USD75,000. Studies since then have confirmed that while both affective and cognitive wellbeing increase with income, income is more strongly related to cognitive- than affective-wellbeing [@diener2013rising; @diener2017findings; @howell2008relation].

Fundamentally, a difference in slope between affective wellbeing and income for the rich and poor represents an unacknowledged source of inequity (i.e., unfairness) in the distribution of wellbeing (i.e., happiness) in the economy. For instance, the "satiety" point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. While much commentary has focused on the zero slope at high income levels reported by Kahneman & Deaton, the positive slope at lower income levels means wellbeing is distributed unequally over those income ranges. Moreover, the location of the change point in slope has implications for the distribution of wellbeing over the entire income range. For instance, an increase in the change point (i.e., a right shift on the income distribution) indicates more inequality in wellbeing between the rich and poor. Conversely, a decrease (left shift) represents a more equal distribution of wellbeing across income levels. To our knowledge, no studies to date have examined whether the location of the change point has shifted over time, but such shifts may help us understand how the distribution of wellbeing has been changing for different parts of the economy.

While studies agree that affective wellbeing tends to increase with income, the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a zero-slope change point after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is approximately log-linear [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon p-value comparisons after dividing the data into arbitrary bins (i.e., deciles, quintiles, quartiles), which can have salient effects on the functional form observed. When p-values have been determined, they have not evaluated the probability of a change point given the data, but rather the probability of the data assuming no change - which is not a direct test.  binning the data (i.e., over income deciles or quintiles) and comparing adjacent bins either visually or with null hypothesis tests which do not provide evidence for a change point. In this regard, no studies to our knowledge have ever directly tested for the presence of a change point between wellbeing and income using standard modelling approaches. For this reason, a formal comparison between log-linear, linear and a change point function (i.e., piecewise) in the relationship between affective wellbeing and income is warranted.

Inequalities in the distribution of wellbeing are increasingly relevant to governments and policy-makers due to the growing recognition that increasing income does not necessarily lead to equal changes in wellbeing [@clark2018four; @frijters2020happy]. Even prior to COVID-19, the World Gallup Poll has observed that affective wellbeing has decreased over the past decade in western Europe, North America, Australia and New Zealand, despite increases in income in the same countries [@sachs2019world]. However to date there has been little investigation of whether the relationship between income and affective wellbeing has changed over time, which may contribute to these trends. In particular, has the functional relationship between income and affective wellbeing, and therefore the distribution of happiness between rich and poor, become more or less equitable in the last few decades?

@@

#### 2.1 Income
We used household after-tax income as the indicator of income and economic security [e..g, @kahneman2010high]. The ‘real household annual disposable income’ was calculated from the self-reported combined income of all household members after receipt of government pensions and benefits and deduction of income taxes in the financial year ended 30th June of the year of the wave (e.g., 2001 in wave 1). This was then adjusted for inflation - the rise in the general price level of the economy - using the Australian Bureau of Statistics (ABS) Consumer Price Index, so that income in all waves is expressed in 2019 prices, to give real income.

@@

We modelled the relationship between income and each wellbeing variable (cognitive and affective) using a linear model, log-linear model, and a piecewise-linear model with a single change point as a free parameter estimated from the data. The piecewise-linear model was chosen as the simplest extension of a linear model which can identify a change point (inflection) in the relationship between wellbeing and income. For modelling, both measures of wellbeing and income were rescaled with a mean of zero and a SD of 1 (z-scores) for each year. As each year was modelled separately this does not affect the results. Importantly, the coefficient for the change point in each year was returned to the original scale and transformed to real 2019 dollars when reported.

@@

_Covariates_  
Covariates for age, age^2^, sex, education, and chronic illness were included in each model above. Cross-sectional population weights for Australia provided by the University of Melbourne for each year were also included as a covariate to adjust for differences in the sample representativeness (by sex, broad age, marital status, region, and labour force status). Thus each model includes terms for the linear effect of each covariate, and age^2^ includes a term for the quadratic effect. Full-time students were removed, as well as individuals with an annual household disposable income that was negative or indicated as _topcoded_ by the University of Melbourne. The results of sensitivity analyses to determine the effect of the covariates and exclusions on the change point parameters are included in Supplementary Materials (Figure S2, S3, S4).

@@

The evidence from model selection revealed the piecewise-linear fit of _affective_- was superior to a log-linear or linear fit in each year with a penalised WAIC score credibly (95% interval) smaller than the log-linear fits for all but three of the nineteen years (and for all 19 years for the linear fits, see Supplementary Materials, Figure S1). There was little evidence that any of the three models were reliably or credibly superior in representing the association between income and cognitive-wellbeing over the same period, albeit the linear model was better on average between 2002-2004 and 2007-2010 while the piecewise-linear model was better the remaining years. Overall the model comparison suggested that affective- and cognitive wellbeing have distinct relationships with household income; affective wellbeing increases with income more rapidly at lower household income levels than higher income levels with a distinct change point in each year, while cognitive wellbeing tends to increase with income, most likely in a linear fashion. Others have also noted cognitive wellbeing (life satisfaction) has a stronger linear relationship with income than affective wellbeing and income [@howell2008relation], and so the results of our formal comparison provides some support for this.

@@

##### Figure 2. Posterior parameters of the affective-wellbeing ~ income piecewise-linear model (real 2019 dollars)   
```{r get_posteriors}
income_params <- read_rds("../results/gh9_hhoecd_summary.RDS")
fits <- read_rds("../results/gh9_hhoecd_0cov_tight.RDS")

posteriors <- map_dfr(fits, as.data.frame, .id = "fit") %>%
  left_join(bind_rows(income_params, .id = "fit")) %>%
  mutate(omega_dollars = (b_omega_Intercept * sd + mean)/1000) %>%
  as_tibble()
```

```{r figure_2, fig.width = 6, fig.height = 7, dpi=300}
param_key = c(
  omega_dollars = "Change point ($000s)",
  b_b0_Intercept = paste0("Intercept (\u03b2", "0)"),
  b_b1_Intercept = paste0("Pre-slope (\u03b2", "1)"),
  b_b2_Intercept = paste0("Post-slope (\u03b2", "2)")
)

posteriors %>%
  select(fit, contains("b0"), contains("b1"), contains("b2"), omega_dollars) %>%
  gather(key = "param", val = "sample", -fit) %>%
  mutate(
    year = extract_numeric(fit),
    param = recode_factor(param, !!!param_key)
    ) %>% 
  group_by(param, year) %>%
  summarise(
    low = quantile(sample, probs = 0.05, na.rm=T),
    expected = quantile(sample, probs = 0.5, na.rm=T),
    upp = quantile(sample, probs = 0.95, na.rm=T)
  ) -> parameter_summary

@@

```
Figure 2 legend: Posterior distributions of the change point parameter representing the location in real household income (real 2019 dollars), as well as the intercept, pre-slope and post-slope parameters in SD units (affective wellbeing). Horizontal bar represents the 95% credible region and the solid point indicates the expected value (median) of each distribution. Vertical dotted line indicates the 2001 expected value (median) as a base year comparison.
```

@@

We found the marginal effect of household income on affective- and cognitive-wellbeing was positive but quite different: Affective wellbeing increased rapidly up to a point after which higher levels of income were associated with less improvement, while cognitive wellbeing tended to increase linearly or constantly with income. Other studies have also reported that income has distinct effects on cognitive and affective wellbeing [@howell2008relation], with a zero-slope change point in affective wellbeing after which increasing income has little further effect [@kahneman2010high]. In our case we did not observe a zero-slope change point ("satiety" point), as the post-slope for each year was credibly above zero (Figure 2); but the change point did represent a substantial decrease in dependency between income and affective wellbeing for those on higher income levels. Our novel finding was revealing the change point increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the change point between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

@@

Other researchers have argued income has a satiety point on affective well-being, after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]. This is consistent with a change point after which the slope is zero. While we observed a change point, the slope afterwards was small but credibly larger than zero so our results are not strictly consistent with a "satiety" point. Our results emphasize the slope over income is not constant. Other researchers have argued that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. We also show the change point has increased over time. Indeed Twenge & Cooper report the linear relationship between income and happiness has become steeper over time in the US General Social Survey (GSS). While we did not see any consistent trend over time in pre or post-slope changes in the piecewise-linear model of affective-wellbeing (e.g., Figure 2), there was a credibly increasing slope in the linear model of cognitive-wellbeing with time (Table S3). It may be the single-item on general happiness in the GSS has more in common with the single item on life-satisfaction in HILDA than the more comprehensive 9-item, dual-valence questions on affective-wellbeing in the SF-36. Nevertheless, consistent with Twenge et al, our results imply that the difference in happiness between the rich and poor has been increasing over time (e.g., Figure 6 and S8). Moreover our piecewise-linear model reveals the income level around which this difference appears.

>>>>

1. Central Clinical School, Faculty of Medicine and Health, University of Sydney, NSW, Australia
2. ARC Centre of Excellence for Children and Families over the Life Course
3. Economic Discipline Group, University of Technology Sydney, NSW, Australia
4. Institute of Labor Economics (IZA), Bonn, Germany

@@

|            |                                     |
|------------|-------------------------------------|
| Draft      | `r format(Sys.time(), '%d %B, %Y')` |
| Words      | 4688                                |
| Tables     | 0                                   |
| Figures    | 4                                   |

@@

A fundamental question for society is how much happiness does a dollar buy? The accepted view among economists and psychologists is that income has diminishing marginal returns on happiness: money and happiness increase together up to a point after which there is relatively little further gain. In this paper we estimate the relationship between income and subjective wellbeing over a 19-year period focusing on where the greatest change in the marginal return on income occurs and whether this change point has shifted over time. We formally test for the presence of a change point as well temporal changes in the relationship between income and affective wellbeing (happiness), and income and cognitive wellbeing (life satisfaction), using household economic data from Australia between 2001-2019. The results indicate that the change point between affective wellbeing and income has increased over those 19 years faster than inflation (i.e., cost of living). This suggests that inequalities in income may be driving increasing _inequities_ in happiness between the rich and the poor, with implications for health and recent government policy-goals to monitor and improve wellbeing.

@@

- We investigate how the relationship between income and subjective wellbeing (happiness and life satisfaction) has changed over time
- Detailed wellbeing data from Australia over an economically stable 19-year period provided an undisturbed view of the changes
- Happiness was less responsive to income after a certain income level, and this level had almost doubled since 2001
- Conversely the relationship between income and life-satisfaction had remained relatively constant over the same period
- A steep association between happiness and income has become more prevalent this millennium, and likely to drive inequities in wellbeing between the rich and poor

@@

@@

Subjective wellbeing is not a unitary entity [@diener2017findings]. Psychology studies typically distinguish between life satisfaction: the cognitive appraisal of one’s own accomplishments; and affective wellbeing: one’s prevailing affective state, emotional mood, or everyday experience of happiness [e.g., @kettlewell2020differential]. While measures of these distinct concepts are correlated, the distinction seems to be critical in understanding the impact of income on wellbeing [@howell2008relation]. For instance, we have recently observed that a major financial gain substantially improved cognitive wellbeing for individuals without much impact on their affective wellbeing [@kettlewell2020differential]. In a now classic paper, Kahneman & Deaton [-@kahneman2010high] showed that affective wellbeing increased with log household income up to a point (USD75,000), but after that the slope between income and wellbeing was effectively zero: increasing income had no further effect (i.e., "income satiation"). Conversely cognitive wellbeing continued to increase with log income beyond USD75,000. Studies since then have confirmed that while both affective and cognitive wellbeing increase with income, (log) income is more strongly related to cognitive than affective wellbeing [@diener2013rising; @diener2017findings; @howell2008relation].

Fundamentally, any difference in slope between affective wellbeing (i.e., happiness) and income for the rich and poor represents an unacknowledged source of inequity (i.e., unfairness) in the distribution of wellbeing in the economy. While much commentary has focused on the zero slope at high income levels reported by Kahneman & Deaton, any steeper slope at lower income levels (relative to high income levels) means wellbeing is distributed more unequally among people with low incomes. For instance, the "satiety" point at USD75,000 in 2008 reported by Kahneman & Deaton was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while that of the richer minority was not.

While studies agree the marginal returns of income on affective wellbeing tend to diminish, the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a zero-slope change point after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]; other studies have found the relationship is approximately log-linear with no discontinuity in slope [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. Rather than assume a log-linear relationship to characterise income and wellbeing, we formally test and compare different functional forms (linear, log-linear and piecewise-linear) to identify the point of greatest change in the marginal effect of income (including but not limited to a zero-slope change); and then determine whether that point is shifting over time. For instance, a right shift of the change point on the income distribution would represent an increase in the income level, and thus indicate more inequality in wellbeing among the poor than the rich. Conversely, a decrease (left shift) would represent a more equal distribution of wellbeing across income levels.

Inequalities in the distribution of wellbeing are increasingly relevant to governments and policy-makers due to the growing recognition that increasing income does not necessarily lead to equal changes in wellbeing [@clark2018four; @frijters2020happy]. Even prior to COVID-19, the World Gallup Poll has observed that affective wellbeing has decreased over the past decade in western Europe, North America, Australia and New Zealand, despite increases in average income in the same countries [@sachs2019world]. However to date there has been little investigation of whether the relationship between income and affective wellbeing has changed over time, which may contribute to these trends. In particular, has the functional relationship between income and affective wellbeing, and therefore the distribution of happiness between rich and poor, become more or less equitable in the last few decades?

@@

#### 2.1 Income
We used household after-tax income as the indicator of income and economic security [e.g., @kahneman2010high]. The ‘real household annual disposable income’ was calculated from the self-reported combined income of all household members after receipt of government pensions and benefits and deduction of income taxes in the financial year ended 30th June of the year of the wave (e.g., 2001 in wave 1). This was then adjusted for inflation - the rise in the general price level of the economy - using the Australian Bureau of Statistics (ABS) Consumer Price Index, so that income in all waves is expressed in 2019 prices, to give real income.

@@

We modelled the relationship between income and each wellbeing variable (cognitive and affective) using a linear model, log-linear model, and a piecewise-linear model with a single change point as a free parameter estimated from the data. The piecewise-linear model was chosen as the simplest extension of a linear model which can identify a change point (discontinuity in slope) in the relationship between wellbeing and income. For modelling, both measures of wellbeing and income were rescaled with a mean of zero and a SD of 1 (z-scores) for each year. As each year was modelled separately this does not affect the results, and the model parameter estimates were returned to their raw score units (e.g., real 2019 dollars) in figures and tables.

@@

_Covariates_  
Covariates for age, age^2^, sex, education, and chronic illness were included in each model above. Cross-sectional population weights for Australia provided by the University of Melbourne for each year were also included as a covariate to adjust for differences in the sample representativeness (by sex, broad age, marital status, region, and labour force status). Thus each model includes terms for the linear effect of each covariate, and age^2^ includes a term for the quadratic effect. Full-time students were removed, as well as individuals with an annual household disposable income that was negative or indicated as _topcoded_ by the University of Melbourne. The results of sensitivity analyses to determine the effect of the covariates and exclusions on the change point parameters are included in Supplementary Materials (Figures S2, S3, and S4).

@@

The evidence from model selection revealed the piecewise-linear fit of _affective_ wellbeing was superior to a log-linear or linear fit in each year with a penalised WAIC score credibly (95% interval) smaller than the log-linear fits for all but three of the nineteen years (and for all 19 years for the linear fits, see Supplementary Materials, Figure S1). There was little evidence that any of the three models were reliably or credibly superior in representing the association between income and cognitive-wellbeing over the same period, albeit the linear model was better on average between 2002-2004 and 2007-2010 while the piecewise-linear model was better the remaining years. Overall the model comparison suggested that affective and cognitive wellbeing have distinct relationships with household income; affective wellbeing increases with income more rapidly at lower household income levels than higher income levels with a distinct change point in each year, while cognitive wellbeing tends to increase with income, most likely in a linear fashion. Others have also noted cognitive wellbeing (life satisfaction) has a stronger linear relationship with income than affective wellbeing and income [@howell2008relation], and so the results of our formal comparison provides some support for this.

@@

##### Figure 2. Posterior parameters of the affective-wellbeing ~ income piecewise-linear model (real 2019 dollars)   
```{r get_posteriors}
# income_params <- read_rds("../results/gh9_hhoecd_summary.RDS")
fits <- read_rds("../results/gh9_hhoecd_0cov_tight.RDS")

# Recover the parameters from standardization to return raw score units
std_params <- list()

for (i in c(2001:2019)) {

  hilda_data %>%
    filter(
      year == i,
      !student,
      hh_disp_inc_eq >= 0,
      !top_hifdip
    ) %>%
    transmute(
      sgh9 = scale(gh9),
      shh_disp_inc_eq = scale(hh_disp_inc_eq),
    ) %>% 
    na.omit() -> df

  income_sd_year <- attr(df$shh_disp_inc_eq, "scaled:scale")
  income_m_year <- attr(df$shh_disp_inc_eq, "scaled:center")
  gh9_sd_year <- attr(df$sgh9, "scaled:scale")
  gh9_m_year <- attr(df$sgh9, "scaled:center")

  std_params[[paste0("fit", i)]] <- c(`mean_income` = income_m_year, 
                       `sd_income` = income_sd_year,
                       `mean_gh9` = gh9_m_year,
                       `sd_gh9` = gh9_sd_year)

}

posteriors <- map_dfr(fits, as.data.frame, .id = "fit") %>%
  left_join(bind_rows(std_params, .id = "fit")) %>%
  transmute(
    fit,
    omega_dollars = (b_omega_Intercept * sd_income + mean_income)/1000,
    b0 = b_b0_Intercept * sd_gh9 + mean_gh9,
    b1 = b_b1_Intercept * sd_gh9,
    b2 = b_b2_Intercept * sd_gh9
    ) %>%
  as_tibble()
```

```{r figure_2, fig.width = 6, fig.height = 7, dpi=300}
param_key = c(
  omega_dollars = "Change point ($000s)",
  b0 = paste0("Intercept (\u03b2", "0)"),
  b1 = paste0("Pre-slope (\u03b2", "1)"),
  b2 = paste0("Post-slope (\u03b2", "2)")
)

posteriors %>%
  gather(key = "param", val = "sample", -fit) %>%
  mutate(
    year = extract_numeric(fit),
    param = recode_factor(param, !!!param_key)
    ) %>% 
  group_by(param, year) %>%
  summarise(
    low = quantile(sample, probs = 0.05, na.rm=T),
    expected = quantile(sample, probs = 0.5, na.rm=T),
    upp = quantile(sample, probs = 0.95, na.rm=T)
  ) -> parameter_summary

@@

```
Figure 2 legend: Posterior distributions of the change point parameter representing the location in real household income (real 2019 dollars), as well as the intercept, pre-slope and post-slope parameters in happiness units (0-100). Horizontal bar represents the 95% credible region and the solid point indicates the expected value (median) of each distribution. Vertical dotted line indicates the 2001 expected value (median) as a base year comparison.
```

@@

We found the marginal effect of household income on affective- and cognitive-wellbeing was positive but quite different: Affective wellbeing increased rapidly up to a point after which higher levels of income were associated with relatively less improvement, while cognitive wellbeing tended to increase linearly with income. Other studies have also reported that income has distinct effects on cognitive and affective wellbeing [@howell2008relation], and our results are consistent with that distinction. We also found a change point in the marginal returns of income on affective wellbeing in each year. Some influential studies have found a zero-slope change point in affective wellbeing after which increasing income had no further effect [@kahneman2010high]. In our case we did not observe a zero-slope change point ("satiety" point), as the post-slope for each year was credibly above zero (Figure 2); but the change point did represent a substantial decrease in dependency between income and affective wellbeing for those on higher income levels. Our novel finding was the change point increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the change point between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

@@

Other researchers have argued income has a satiety point on affective well-being, after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]. This is consistent with a change point after which the slope is zero. While we observed a change point, the slope afterwards was small but credibly larger than zero so our results are not strictly consistent with a "satiety" point. Our results emphasize the slope over income is not constant. Other researchers have also argued that affective wellbeing is a continuous log-linear function of income [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding], which our results do not support and instead suggest there is a point (at least one) of discontinuity in the change of slope over income. We also show the change point has increased over time. Indeed Twenge & Cooper report the linear relationship between income and happiness has become steeper over time in the US General Social Survey (GSS). While we did not see any consistent trend over time in pre or post-slope changes in the piecewise-linear model of affective-wellbeing (e.g., Figure 2), there was a credibly increasing slope in the linear model of cognitive-wellbeing with time (Table S3). It may be the single-item on general happiness in the GSS has more in common with the single item on life-satisfaction in HILDA than the more comprehensive 9-item, dual-valence questions on affective-wellbeing in the SF-36. Nevertheless, consistent with Twenge et al, our results imply that the difference in happiness between the rich and poor has been increasing over time (e.g., Figure 6 and S8). Moreover our model reveals the income level around which the greatest difference appears.
